---
title: "Chapter 4 Exercise - Time series features"
author: "Matt Colantonio"
date: "`r Sys.Date()`"
output: 
  html_document:
  toc: true
  toc_float: true
  theme: lumen
---

```{r message=FALSE, warning=FALSE, include=FALSE}
# this code clears the environment and installs/loads popular packages

rm(list = ls()) 
  gc()            
  cat("\f")  
packages <- c("readr", #open csv
              "psych", # quick summary stats for data exploration,
              "stargazer", #summary stats for sharing,
              "tidyverse", # data manipulation like selecting variables,
              "corrplot", # correlation plots
              "ggplot2", # graphing
              "GGally", # ts features
              "ggcorrplot", # correlation plot
              "gridExtra", #overlay plots
              "data.table", # reshape for graphing 
              "car", #vif
              "prettydoc", # html output
              "visdat", # visualize missing variables
              "glmnet", # lasso/ridge
              "caret", # confusion matrix
              "MASS", #step AIC
              "plm", # fixed effects demeaned regression
              "lmtest", # test regression coefficients
              "fpp3", # Foprecasting: Principles & Practice supplement
              "tsibble", 
              "tsibbledata",
              "lubridate",
              "forecast"
)

for (i in 1:length(packages)) {
  if (!packages[i] %in% rownames(installed.packages())) {
    install.packages(packages[i]
                     , repos = "http://cran.rstudio.com/"
                     , dependencies = TRUE
    )
  }
  library(packages[i], character.only = TRUE)
}

rm(packages)
  
```

```{r include=FALSE}
setwd("/Users/matthewcolantonio/Documents/Textbooks/Forecasting-Hyndman/")
```

#### 1. Write a function to compute the mean and standard deviation of a time series, and apply it to the `PBS` data. Plot the series with the highest mean, and the series with the lowest standard deviation.

```{r}
df <- PBS

compute_mean_sd <- function(x) {
  mean_value <- mean(x)
  sd_value <- sd(x)
  return(list(mean=mean_value, sd=sd_value))
}
```

```{r}
results <- lapply(df, compute_mean_sd)

```

```{r}
highest_mean_series <- names(results)[which.max(sapply(results, function(x) x$mean))]
lowest_sd_series <- names(results)[which.min(sapply(results, function(x) x$sd))]


```

Highest mean is Cost.

```{r}
PBS %>% autoplot(Cost, show.legend = F)
```

Lowest sd is Month (this seems quite obvious).

```{r}
PBS %>% autoplot(Month, show.legend = F)
```

#### 2. Use `GGally::ggpairs()` to look at the relationships between the STL-based features for the holiday series in the `tourism` data. Change `seasonal_peak_year` and `seasonal_trough_year` to factors, as shown in Figure [4.3](https://otexts.com/fpp3/exploring-australian-tourism-data.html#fig:seasonalfeatures). Which is the peak quarter for holidays in each state?

```{r}
data("tourism")
holiday_data <- filter(tourism, Purpose== "Holiday")

```

For Adelaide:

```{r}
holiday_data[holiday_data$Region == "Adelaide", ] %>%
  model(
    STL(Trips ~ trend(window = 7) + 
          season(window = "periodic"),
        robust = TRUE)) %>%
  components() %>%
  autoplot()

```

For all Regions:

```{r}
holiday_data %>%
  model(
    STL(Trips ~ trend(window = 7) + 
          season(window = "periodic"),
        robust = TRUE)) %>%
  components() %>%
  autoplot(show.legend = FALSE)
```

Creating objects out of the STL models so that the seasonal aspects can be made into factors and compared.

```{r}
stl <- holiday_data %>%
  model(
    STL(Trips ~ trend(window = 7) + 
          season(window = "periodic"),
        robust = TRUE)) %>%
  components() %>%
  autoplot(show.legend = FALSE)
```

#### 3. Use a feature-based approach to look for outlying series in the `PBS` data. What is unusual about the series you identify as "outliers".

```{r}
feature_df <- df %>%
  index_by(Month) %>%
  summarise(
    Mean_Scripts = mean(Scripts),
    StdDev_Scripts = sd(Scripts),
    Mean_Cost = mean(Cost),
    StdDev_Cost = sd(Cost)
    # Add more features as needed
  )

# Determine outliers based on the features
# For example, you can use a z-score threshold to identify outliers
z_threshold <- 1.96  # Adjust this threshold as needed
outliers <- feature_df %>%
  filter(abs(Mean_Scripts) > z_threshold | abs(StdDev_Scripts) > z_threshold | abs(Mean_Cost) > z_threshold | abs(StdDev_Cost) > z_threshold)

# Print the outliers
print(outliers)
```

This list of outliers is 204 observations long, using Z-score of 1.96. There are 6080 observations in the holiday dataset.

The GGally plot is also useful and it looks cool.

```{r}
holiday_features <- holiday_data %>%
  features(Trips, feature_set(pkgs = "feasts"))
holiday_features


```

```{r message=FALSE, warning=FALSE}
library(glue)
holiday_features %>%
  select_at(vars(contains("season"), Purpose)) %>%
  mutate(
    seasonal_peak_year = seasonal_peak_year +
      4*(seasonal_peak_year==0),
    seasonal_trough_year = seasonal_trough_year +
      4*(seasonal_trough_year==0),
    seasonal_peak_year = glue("Q{seasonal_peak_year}"),
    seasonal_trough_year = glue("Q{seasonal_trough_year}"),
  ) %>%
  GGally::ggpairs(mapping = aes(colour = Purpose))
```
